---
title: "SqueakR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SqueakR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The SqueakR package is a centralized, experiment-centered approach to organizing and analyzing data exported from [`DeepSqueak`](https://github.com/DrCoffey/DeepSqueak). Of its diverse functions, SqueakR is capable of generating quick plots from vocalization data detected from `DeepSqueak`. On top of these visualizations, the package contains functions which combine all exported DeepSqueak data from multiple recordings into a single experimental object, allowing subsequent analysis to be conducted directly from R.

# Installation

## From GitHub

To use SqueakR, we need to install the package. To do so, run the code below. Ensure the `devtools` package is installed before running the following:

```{r install, eval=FALSE}
library(devtools)
install_github("osimon81/SqueakR")
```

Next, load the package:

```{r setup}
library(SqueakR)
```


# The Experimental Object

This package allows all data necessary for visualizations to be stored in a single object. However, to do so, the experimental object must be created. To create an `experiment`, run the following code:

```{r create experiment}
experiment <- create_experiment(experiment_name = "my_experiment")
```

This code creates a new `experiment` object, with the name `my_experiment`. When the object is saved, it will be saved using this name, along with a timestamp. At this point, there's nothing stored in this `experiment` object except its name, but let's inspect how `experiment` is structured:

```{r}
str(experiment)
```

We can see from the `str()` function that the `experiment` object has 5 main groups listed in it:
- name: The name we just set for this experiment
- last_saved: A timestamp for the last time this experiment was saved (in this case, this is the time the object was created)
- groups: An empty variable which will be configured to show the experimental groups
- experimenters: An empty variable which will be configured to show the experimenters who collected data
- experimental_data: An empty list which will store all of the data for this experiment

## Creating an Experiment






